---
title: "Wildfire Network Analysis"
author: "Atharva Jagtap, Ishika Agarwal, Mariya Putwa, Prateek Balani"
date: "2024-11-11"
output: pdf_document
---
# Big network plot

```{r}
# Load necessary libraries
library(dplyr)
library(igraph)
library(geosphere)
library(ggplot2)
library(sf)
library(RColorBrewer)
library(viridis)
library(stringr)
library(dbscan)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)


wildfire_data <- read.csv("data_clean.csv")

# Filter top 100 fires per year based on `current_size`
top_fires <- wildfire_data %>%
  group_by(fire_year) %>%
  arrange(desc(current_size)) %>%
  slice_head(n = 25) %>%
  ungroup()

# Create a unique identifier by combining `fire_year` and `fire_number`
top_fires <- top_fires %>%
  mutate(unique_id = paste(fire_year, fire_number, sep = "_"))

# Prepare nodes (distinct fires with geographical coordinates)
nodes <- top_fires %>%
  select(unique_id, fire_location_latitude, fire_location_longitude, current_size, fire_year, activity_class) %>%
  distinct()

# Define edges based on geographical proximity within each year
max_distance <- 50000  # Set max distance to 20 km for connecting fires
# Define edges based on geographical proximity within each `activity_class`
edges <- data.frame(from = character(), to = character(), stringsAsFactors = FALSE)

# Loop to create edges within each `activity_class` group based on proximity
for (class in unique(top_fires$activity_class)) {
  # Filter data for the current activity class and remove rows with missing coordinates
  class_data <- top_fires %>%
    filter(activity_class == class) %>%
    filter(!is.na(fire_location_latitude) & !is.na(fire_location_longitude))
  
  # Skip if no rows remain after filtering
  if (nrow(class_data) < 2) {
    next  # Skip this iteration if there aren’t enough points to form edges
  }
  
  # Nested loop to calculate distances and create edges
  for (i in 1:(nrow(class_data) - 1)) {
    for (j in (i + 1):nrow(class_data)) {
      # Extract the coordinates as pairs
      coord_i <- c(class_data$fire_location_longitude[i], class_data$fire_location_latitude[i])
      coord_j <- c(class_data$fire_location_longitude[j], class_data$fire_location_latitude[j])
      
      # Check if coordinates are valid pairs of length 2
      if (length(coord_i) == 2 && length(coord_j) == 2) {
        # Calculate distance between points
        dist <- distHaversine(coord_i, coord_j)
        
        # Add edge if distance is within the maximum allowed distance
        if (!is.na(dist) && dist <= max_distance) {
          edges <- rbind(edges, data.frame(from = class_data$unique_id[i], to = class_data$unique_id[j]))
        }
      }
    }
  }
}


# Filter edges to ensure IDs are in nodes
edges <- edges %>%
  filter(from %in% nodes$unique_id & to %in% nodes$unique_id)

# Create the graph from edges and nodes
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

V(g)$activity_class <- nodes$activity_class

# Map `activity_class` to colors
activity_classes <- unique(nodes$activity_class)
color_palette <- brewer.pal(n = length(activity_classes), name = "Set3")  # Customize with any color palette you prefer
activity_colors <- setNames(color_palette, activity_classes)  # Map each activity_class to a color


# Apply colors to nodes based on `activity_class`
V(g)$color <- activity_colors[V(g)$activity_class]

# Node sizes based on `current_size`
V(g)$size <- log(nodes$current_size + 1) * 2

# Adjust node colors for transparency
node_colors <- adjustcolor(V(g)$color, alpha.f = 0.5)  # 70% opacity

# Adjust edge color for transparency
edge_color <- adjustcolor("red", alpha.f = 1) 

# Plot the network with Kamada-Kawai layout
layout_kk <- layout_with_kk(g)
plot(g,
     layout = layout_kk,
     vertex.size = V(g)$size,
     vertex.color = node_colors,  # Transparent node colors
     vertex.label = NA,
     edge.color = edge_color,      # Transparent edge color
     edge.width = 2,
     main = "Wildfire Network of Top 25 Fires per Year KK (Clustered by Activity Class)")

legend("right", 
       legend = names(activity_colors),
       col = activity_colors,
       pch = 16,
       pt.cex = 1.5,   # Increase point size in legend for visibility
       title = "Activity Class",
       xpd = TRUE,                
       inset = -0.05,
       bg = adjustcolor("white", alpha.f = 0.8))  # Transparent legend background

layout_fr <- layout_with_fr(g)
plot(g,
     layout = layout_fr,
     vertex.size = V(g)$size,
     vertex.color = node_colors,
     vertex.label = NA,
     edge.color = edge_color,
     edge.width = 2,   # Keep a higher width for better visibility
     main = "Wildfire Network of Top 25 Fires per Year FR (Clustered by Activity Class)")

legend("right", 
       legend = names(activity_colors),
       col = activity_colors,
       pch = 16,
       pt.cex = 1.5,   # Increase point size in legend for visibility
       title = "Activity Class",
       xpd = TRUE,                
       inset = -0.05,
       bg = adjustcolor("white", alpha.f = 0.8))  # Transparent legend background

cat("Number of edges in graph:", ecount(g), "\n")
cat("Number of vertices in graph:", vcount(g), "\n")
```

# plot for each year
```{r}
# Fixed bounding box for Canada
register_stadiamaps("59b82ef6-c67a-4d54-ac10-301e0edb43e4", write = TRUE)

bbox <- c(left = -140, bottom = 40, right = -50, top = 85)  # Covers Canada

# Convert nodes data to an sf object for spatial plotting
nodes_sf <- st_as_sf(nodes, coords = c("fire_location_longitude", "fire_location_latitude"), crs = 4326)

# Reproject to Web Mercator (EPSG:3857) to match Stamen map tiles
nodes_sf <- st_transform(nodes_sf, crs = 3857)

# Loop to create a map for each year
years <- unique(top_fires$fire_year)

for (year in years) {
  yearly_data <- nodes_sf %>% filter(fire_year == year)
  
  # Get a background map centered on Canada
map_base <- get_stadiamap(
    bbox = bbox,   # Use bounding box for each year's data
    zoom = 4,      # Adjust zoom level if needed
    maptype = "outdoors"  # Options: "outdoors", "alidade_smooth_dark"
  )

  # Plot the map with wildfire clusters overlaid
  ggmap(map_base) +
    geom_sf(data = yearly_data, aes(color = as.factor(cluster), size = log(current_size + 1)), inherit.aes = FALSE) +
    scale_color_viridis_d(option = "plasma") +
    labs(title = paste("Wildfire Locations and Clusters for Year:", year),
         color = "Cluster") +
    theme_minimal() +
    theme(legend.position = "bottom") +
    scale_size(range = c(2, 6))
}
```

```{r}
# Data Analysis on the given data 
Degree_centrality <- degree(g)
Closeness_centrality <- closeness(g)
Betweenness_centrality <- betweenness(g)
Eigenvector_centrality <- evcent(g)$vector
Clustering_coefficient <- transitivity(g, type = "local")

# Combine centrality measures into a data frame
centrality_measures <- data.frame(
  node = V(g)$name,
  degree = Degree_centrality,
  closeness = Closeness_centrality,
  betweenness = Betweenness_centrality,
  eigenvector = Eigenvector_centrality,
  clustering = Clustering_coefficient
)

summary(centrality_measures)

# Top 5 nodes by degree centrality
top_degree <- centrality_measures[order(-centrality_measures$degree), ]
head(top_degree, 5)

# Top 5 nodes by closeness centrality
top_closeness <- centrality_measures[order(-centrality_measures$closeness), ]
head(top_closeness, 5)

# Top 5 nodes by betweenness centrality
top_betweenness <- centrality_measures[order(-centrality_measures$betweenness), ]
head(top_betweenness, 5)

# Top 5 nodes by eigenvector centrality
top_eigenvector <- centrality_measures[order(-centrality_measures$eigenvector), ]
head(top_eigenvector, 5)

# Top 5 nodes by clustering coefficient
top_clustering <- centrality_measures[order(-centrality_measures$clustering), ]
head(top_clustering, 5)


# Plotting the centrality measures




```

```{r}

# Common causes of wildfires 


#filtering data as to prevent any missing values and in turn creating issues 

wildfire_filtered <- wildfire_data %>%
  mutate(across(where(is.character), ~ trimws(.))) %>% # Remove whitespace
  filter(!is.na(fire_location_latitude), !is.na(fire_location_longitude), !is.na(activity_class)) %>%
  mutate(activity_class = as.factor(activity_class))


# Converting wildire coordinates to spatial data and using cors=4326 as it is the most common coordinate system for latitude and longitude.
wildfire_sf <- st_as_sf(
  wildfire_filtered,
  coords = c("fire_location_longitude", "fire_location_latitude"),
  crs = 4326
)

#matrix that stores a point (i.e wildire ) and its coordinates
coords <- st_coordinates(wildfire_sf)


wild_fire_db <- dbscan(coords, eps = 0.2, minPts = 5) #using a small eps value to prevent any overlapping of clusters and having a better understanding of the causes of the fire 

# Add cluster information to the spatial data
wildfire_sf <- wildfire_sf %>%
  mutate(cluster = as.factor(wild_fire_db$cluster))


# Finding the common causes of wildfires and their average size
common_causes <- wildfire_sf %>%
  group_by(activity_class) %>%
  summarise(
    count = n(),
    avg_fire_size = mean(current_size, na.rm = TRUE)
  ) %>%
  arrange(desc(count))

print("Most Common Causes of Wildfires:")
print(common_causes)

# Effect of causes on spread and clustering
spread_and_clustering <- wildfire_sf %>%
  group_by(activity_class) %>%
  summarise(
    avg_spread_rate = mean(fire_spread_rate, na.rm = TRUE),
    avg_cluster_size = mean(current_size, na.rm = TRUE),
    cluster_count = n_distinct(cluster)
  )%>%
  arrange(desc(cluster_count))

print("Effect of Causes on Spread and Clustering:")
print(spread_and_clustering)

# Plotting the wildfire clusters on a map of Canada
canada_map <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(admin == "Canada")

# Plot wildfire clusters on the map
ggplot() +
  geom_sf(data = canada_map, fill = "lightgray", color = "white") + # Map of Canada
  geom_sf(data = wildfire_sf, aes(color = activity_class), alpha = 0.8) + # Wildfire points
  labs(
    title = "Wildfire Clusters in Canada",
    color = "Activity Class"
  ) +
  theme_minimal()


```


# Calculate network density
network_density <- edge_density(g)
cat("\nNetwork Density:", network_density)
```
```{r}
# 1. Visualize Degree Centrality Distribution
degree_centrality_df <- data.frame(node = names(Degree_centrality), degree = Degree_centrality)
ggplot(degree_centrality_df, aes(x = degree)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Degree Centrality Distribution", x = "Degree Centrality", y = "Frequency")

# 2. Visualize Closeness Centrality Distribution
closeness_centrality_df <- data.frame(node = names(Closeness_centrality), closeness = Closeness_centrality)
ggplot(closeness_centrality_df, aes(x = closeness)) +
  geom_histogram(bins = 30, fill = "salmon", color = "black") +
  labs(title = "Closeness Centrality Distribution", x = "Closeness Centrality", y = "Frequency")

# 3. Visualize Betweenness Centrality Distribution
betweenness_centrality_df <- data.frame(node = names(Betweenness_centrality), betweenness = Betweenness_centrality)
ggplot(betweenness_centrality_df, aes(x = betweenness)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Betweenness Centrality Distribution", x = "Betweenness Centrality", y = "Frequency")

# 4. Visualize Eigenvector Centrality Distribution
eigenvector_centrality_df <- data.frame(node = names(Eigenvector_centrality), eigenvector = Eigenvector_centrality)
ggplot(eigenvector_centrality_df, aes(x = eigenvector)) +
  geom_histogram(bins = 30, fill = "purple", color = "black") +
  labs(title = "Eigenvector Centrality Distribution", x = "Eigenvector Centrality", y = "Frequency")

# 5. Community Detection and Visualization
# Detect communities using a community detection algorithm (Louvain)
communities <- cluster_louvain(g)
V(g)$community <- membership(communities)

# Count the total number of communities
total_communities <- length(unique(V(g)$community))
cat("Total number of communities:", total_communities)

# Identify the largest community and its size
community_sizes <- sizes(communities)
largest_community <- which.max(community_sizes)
largest_community_size <- community_sizes[largest_community]
cat("\nSize of the largest community:", largest_community_size)

# Plot the network with nodes
set.seed(42)  # For reproducible layout
par(mar = c(1, 1, 1, 1))
plot(g, 
     vertex.color = adjustcolor(V(g)$color, alpha.f = 0.5),
     vertex.size = V(g)$size,
     vertex.label = NA,
     edge.width = 1.5,
     edge.color = adjustcolor("red"),
     main = "Wildfire Network with Communities Highlighted")
```

